On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-10-04 Manuscript/manuscript.tex
edits/comments

%Abstract should be 200 words
%Devon Allen was disqualified in 2022... discussions about dq, talk methods and 
%models, venue effect, and conclusion
\begin{abstract}
Devon Allen was disqualified in the men's 110 meter hurdle final of the 2022
World Track and Field Championships after registering a reaction time of 0.099 
seconds, 0.001 seconds faster than what is allowed. Following the games, 
bloggers on the running website LetsRun concluded that the reaction times  
from the 2022 World Championships seemed to be generally faster compared to the 
other reaction times they considered, but they did not perform any formal 
statistical analysis. This paper questions if athletes who competed at the 2022
World Track and Field Championship had similar times at other competitions and
if 0.1 seconds is a reasonable disqualification barrier. To do so, we employ a 
signed-rank test for clustered data to compare reaction times for the same 
athletes at different competitions and a generalized linear mixed model 
(GLMM) with random venue (year) effect and a random heat effect within venue 
in order to model the reaction time data. This matter needs to be addressed 
because disqualification based on allowable reaction time will continue to be an
issue in future world championships.

@@

With our small sample size, the asymptotic normal distribution may not be
reliable, so we also use 1000 random permutations to simulate the null 
distribution of the test statistic.
This method is available from the \texttt{clusWilcox.test()} function
with \texttt{method = `ds'} (for \underline{D}atta and \underline{S}atten) and 
\texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

@@

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $\nu = ab^2$. The
variance as a function of the mean~$\mu$ in the GLM sense is
$\text{var}(\mu) = \mu^2 / a$, which implies that the dispersion parameter is
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ijk}$ given the venue
effect~$v_i$ and heat effect~$h_{i/j}$
has shape $a = 1 / \phi$ and scale
\jy{Note that, as used in the code, we should have
  \texttt{sigma(model) \^ 2 =  phi}. So in Table 2, we can report phi in the last column.}
$b = g^{-1}(\alpha + v_i + h_{i/j}) \phi$. The marginal
distribution of $Y_{ijk}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. Many
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We are specifically interested in the probability of a reaction time
being less than 0.1 seconds in order to gauge if that is a reasonable 
disqualification barrier.

@@

Table~\ref{tab:Clusrankresults} shows the results generated from both the 
permutation (exact) and asymptotic rank-based tests for the three comparisons, 
as discussed previously in Section~\ref{sec:rank}. The national versus international
comparison produced lower p-values than the 2019 versus 2022 comparison, with the
national vs international permutation test producing a p-value of $2\cdot10^{-6}$.
This is a highly significant test result and shows that athletes who competed at the
2022 World Track and Field Championships were on average, significantly faster
at reacting than they were just a few months prior.  For the 2019 versus 2022
same athlete comparison with permutation test, the p-value was 
$2.9\cdot10^{-3}$, which is significant at $\alpha = 0.01$.  The result of
comparing 2022 and 2023 was very similar to the above result, as the p-value
from the permutation test was $3.0\cdot10^{-3}$.  Taken together, these three
tests help show that athletes competing at the 2022 World Track and Field
Championships had faster reaction times than when they had competed previously.

@@

The estimated parameters of the GLMM model from our analysis are shown in 
Table~\ref{tab:Gamma_parameters}.  These statistics reflect the inclusion of 2022
in our analysis and shows the three models we considered. $\sigma_h^2$ is the
standard deviation of the heat effect, $\sigma_v^2$ is the standard deviation of
the venue effect, $\alpha$ is the intercept, and $a$ is the shape parameter of
the gamma function.
The GLMM with both venue and heat random
effects was the best model as evidenced by a decrease in the AIC compared to the
other models with only one random effect. 
Thus, results from the GLMM with both venue and heat random effects
will be presented throughout the rest
of the paper.

\begin{table}
  \centering
  \caption{Estimated parameters for the two effect GLMM on various data sets. SE
  is short for standard error. 
	\eds{This table is not showing what you have described here or in the
          main text.}
        \jy{report $ a = 1 / \phi$}
      } 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $a$ \\
      \midrule
      Venue Effect Only & $-2023.5$ & N/A & 0.054 & $-1.871$ & 0.150\\
      Heat Effect Only & $-2070.7$ & 0.073 & N/A & $-1.890$ & 0.135\\
      Venue and Heat Effect & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

@@

\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than 0.08, 0.09, and
  0.01 seconds based on the two effect model.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & P(R. Time $<$ 0.08) & P(R. Time $<$ 0.09) & P(R. Time $<$ 0.1)  \\ 
   \midrule
   Excluding 2022 & $3.35\cdot10^{-5}$ & $4.45\cdot10^{-4}$ &  $3.54\cdot10^{-3}$  \\ 
   Including 2022 & $4.20\cdot10^{-5}$ & $5.70\cdot10^{-4}$ & $4.34\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

Table~\ref{tab:Sim_probability} shows that changing the reaction time 
barrier from 0.1 seconds to 0.08 seconds drastically reduces the chances of 
observing a reaction time that would break the barrier. If World Athletics were
to change the reaction time the way that \citet{komi2009iaaf} suggests, the 
probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification.

@@

\begin{table}
  \centering
  \caption{Suggested reaction time barriers based on probability of observing
  a time less than the barrier. \of{Not sure if this makes sense}}
  \begin{tabular}{c c c} 
   \toprule
   Data Set & P(RxnTime $< x) = 10^{-3}$ & P(RxnTime $< x) = 10^{-4}$ \\ 
   \midrule
   Excluding 2022 & $0.094$ & $0.084$ \\ 
   Including 2022 & $0.092$ & $0.083$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_time}
\end{table}

@@

\bibliographystyle{chicago}
\bibliography{citations.bib}

>>>>

%Abstract should be 200 words
%Devon Allen was disqualified in 2022... discussions about dq, talk methods and 
%models, venue effect, and conclusion
\begin{abstract}
Devon Allen was disqualified in the men's 110 meter hurdle final of the 2022
World Track and Field Championships after registering a reaction time of 0.099 
seconds, 0.001 seconds faster than what is allowed. Following the games, 
bloggers on the running website LetsRun concluded that the reaction times  
from the 2022 World Championships seemed to be generally faster compared to the 
other reaction times they considered, but they did not perform any formal 
statistical analysis. This paper questions if athletes who competed at the 2022
World Track and Field Championship had similar times at other competitions and
if 0.1 seconds is a reasonable disqualification barrier. To do so, we employ a 
signed-rank test for clustered data to compare reaction times for the same 
athletes at different competitions and a generalized linear mixed model 
with random venue (year) effect and a random heat effect within venue 
in order to model the reaction time data. This matter needs to be addressed 
because disqualification based on allowable reaction time will continue to be an
issue in future world championships.

@@

With our small sample size, the asymptotic normal distribution may not be
reliable, so we also use 1~million random permutations to simulate the null 
distribution of the test statistic.
This method is available from the \texttt{clusWilcox.test()} function
with \texttt{method = `ds'} (for \underline{D}atta and \underline{S}atten) and 
\texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

@@

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $\nu = ab^2$. The
variance as a function of the mean~$\mu$ in the GLM sense is
$\text{var}(\mu) = \mu^2 / a$, which implies that the dispersion parameter is
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ijk}$ given the venue
effect~$v_i$ and heat effect~$h_{i/j}$
has shape $a = 1 / \phi$ and scale
$b = g^{-1}(\alpha + v_i + h_{i/j}) \phi$. The marginal
distribution of $Y_{ijk}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. Many
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We are specifically interested in the probability of a reaction time
being less than 0.1 seconds in order to gauge if that is a reasonable 
disqualification barrier.

@@

Table~\ref{tab:Clusrankresults} shows the results generated from both the 
permutation and asymptotic rank-based tests for the three comparisons, 
as discussed previously in Section~\ref{sec:rank}. The national versus international
comparison produced lower p-values than the 2019 versus 2022 comparison, with the
national vs international permutation test producing a p-value of $2\cdot10^{-6}$.
This is a highly significant test result and shows that athletes who competed at the
2022 World Track and Field Championships were on average, significantly faster
at reacting than they were just a few months prior.  For the 2019 versus 2022
same athlete comparison with permutation test, the p-value was 
$2.9\cdot10^{-3}$, which is significant at $\alpha = 0.01$.  The result of
comparing 2022 and 2023 was very similar to the above result, as the p-value
from the permutation test was $3.0\cdot10^{-3}$.  Taken together, these three
tests help show that athletes competing at the 2022 World Track and Field
Championships had faster reaction times than when they had competed previously.

@@

The estimated parameters of the GLMM model from our analysis are shown in 
Table~\ref{tab:Gamma_parameters}.  These statistics reflect the inclusion of 2022
in our analysis and shows the three models we considered. $\sigma_h^2$ is the
\jy{read what you wrote. These are variances not sd. Do not start a sentence
  with a symbol (see my writing guidelines).}
standard deviation of the heat effect, $\sigma_v^2$ is the standard deviation of
the venue effect, $\alpha$ is the intercept, and $\phi$ is the dispersion
parameter of the gamma model.
The GLMM with both venue and heat random
effects was the best model as evidenced by a decrease in the AIC compared to the
other models with only one random effect. 
Thus, results from the GLMM with both venue and heat random effects
will be presented throughout the rest
of the paper.

\jy{do a thorough check on the captions of tables and figures.}

\begin{table}
  \centering
  \caption{AIC and estimated parameters for the three GLMMs fitted to the ???
    data.}
  \label{tab:Gamma_parameters}
  \begin{tabular}{c c c c c c}
      \toprule
      Model & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\sqrt{\phi}$ \\
      \midrule
      Venue Effect Only & $-2023.5$ &   & 0.054 & $-1.871$ & 0.150\\
      Heat Effect Only & $-2070.7$ & 0.073 &  & $-1.890$ & 0.135\\
      Venue and Heat Effect & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
      \bottomrule
  \end{tabular}
\end{table}

@@

\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than threshold 0.08, 0.09, and
  0.10 seconds based on the two effect model.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & Threshold 0.08 & Threshold 0.09 & Threshold 0.10  \\ 
   \midrule
   Excluding 2022 & $3.35\cdot10^{-5}$ & $4.45\cdot10^{-4}$ &  $3.54\cdot10^{-3}$  \\ 
   Including 2022 & $4.20\cdot10^{-5}$ & $5.70\cdot10^{-4}$ & $4.34\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

Table~\ref{tab:Sim_probability} shows that changing the reaction time 
barrier from 0.1 seconds to 0.08 seconds drastically reduces the chances of 
observing a reaction time that would break the barrier. If World Athletics were
to change the reaction time the way that \citet{komi2009iaaf} suggests, the 
probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification.
\jy{discuss the differences between with/without 2022 data; comment on the
  pecularity of the 2022 data (as suggested by the rank based method); In a
  longer run, it may not be as strange as it looks at the first time.}

\begin{table}
  \centering
  \caption{Suggested reaction time barriers based on tail probabilities.}
  \begin{tabular}{c c c} 
   \toprule
   Data Set & Tail probability  $10^{-3}$ & Tail probability $10^{-4}$ \\ 
   \midrule
   Excluding 2022 & $0.094$ & $0.084$ \\ 
   Including 2022 & $0.092$ & $0.083$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_time}
\end{table}

\jy{be consistent in using scienfit format of numbers; 0.0001 vs $10^{-4}$}

@@

@@

\bibliographystyle{chicago}
\bibliography{citations}
